Mining the human microbiome for clinical insight

### 2018-12-20 24hr/24hr_ms.tex
Fix missing refs

Wastewater-based epidemiology has been proposed for many exciting applications. It has been shown to be useful as an early indicator of poliovirus outbreaks to target vaccination campaigns in polio-free countries \cite{Manor2014,Kaliner2015} and to quantify asymptomatic circulation \cite{Berchenko2017}. Environmental surveillance of typhoid and antimicrobial resistance through urban wastewater are emerging areas of interest in global and public health organizations \cite{Pehrsson2017, Initiatives2018}. Besides infectious disease, methods have been developed and standardized over the last decade to successfully provide continuous data on illicit drug consumption at the national- and regional-level in Australia \cite{AustralianReport} and over 60 European cities and towns \cite{vanNuijs2018,Thomas2012,BazLomba2016}. The academic community is now interested in branching out to measure lifestyle chemicals and biomarkers of health such as obesity levels \cite{Daughton2018,Arnaud2018,Newton2015}. More recently, wastewater testing has been proposed as a promising tool to tackle the opioid epidemic that is severely affecting the U.S \cite{Keshaviah2016,Keshaviah2017}. However, apart from successful poliovirus surveillance and, to a lesser extent, national-level drug consumption surveillance programs, most application areas remain proof-of-concept academic projects and have yet to be implemented into public health and other public service delivery agencies.

@@

Besides providing geographical specificity to the sample, residential catchments mitigate variability in population size and sewage travel time observed at WWTPs. Our selection criteria produces catchments with similar population sizes by design (5,000--10,000 inhabitants) rather than the large range observed (1,000--2,000,000 inhabitants) in WWTPs (Supp. Fig. \ref{24hr:figS1}, \cite{Newtown2015}). Additionally, residential catchments are more likely to contain even representation of all households in the catchment compared to WWTPs: in our residential catchment, sewage from every source travels at most 45 minutes to the sampling point and the distribution of travel times is narrowly-distributed (Fig. \ref{24hr:fig1}A). Sewage at WWTPs, on the other hand, can have travelled anywhere from minutes to hours before it reaches the sampling point. Sampling in residential catchments therefore mitigates the need to correct for unequal population sizes and travel times to some extent.

@@

Metabolites and bacteria in sewage separate into human and environmental contributions. Groups M1 and O1 were enriched in human-derived bacteria and metabolites, while groups M2 and O2 contained bacteria and metabolites derived from the environment (Fig. \ref{24hr:fig3}C). The cluster M1 was statistically enriched in matches to human metabolites compared to the M2 cluster (Fig. \ref{24hr:fig3}C), and includes known urinary and fecal metabolites. Twenty-two metabolic features in M1 were identified as glucuronide compounds of hormones, bile acids and pharmaceuticals, which are direct markers of human excretion as discussed above. Bacteria in the cluster O1 were putatively identified as coming from human gut microbiomes. Cluster O1 included 41 bacterial OTUs primarily from the Firmicutes and Bacteroidetes phyla, which are major components of the human gut flora (Fig. \ref24hr:fig3}D). Additionally, the top 10 most abundant bacterial families in human stool (Human Microbiome Project) made up 65\% of the community in the O1 group (Supp. Fig. \ref{24hr:figS4}). Similar to what was observed in the metabolites, O1 bacteria had a dip in relative abundance during nighttime (Figure \ref{24hr:fig3}B). Clusters M2 and O2, on the other hand, remained relatively constant through the day and likely reflect bacteria and metabolites sourced from environmental background rather than human activity. Cluster M2 had few putative human-associated hits, and had constant abundance through the day and nighttime. The molecules in this group likely represent chemicals derived from the environment or sewer biochemistry. Human-associated bacteria were similarly decreased in cluster O2: most of the OTUs were Proteobacteria, which are not major members of the normal gut flora, and human fecal bacterial families comprised only 1.4\% of the abundance of this cluster (Supp. Fig. \ref{24hr:figS4}). The relative abundance of cluster O2 increased at night, reflecting that these environmental made up a larger proportion of the residential sewage microbiome during periods of low human activity (Fig. \ref{24hr:fig3}D).

>>>>

Wastewater-based epidemiology has been proposed for many exciting applications. It has been shown to be useful as an early indicator of poliovirus outbreaks to target vaccination campaigns in polio-free countries \cite{Manor2014,Kaliner2015} and to quantify asymptomatic circulation \cite{Berchenko2017}. Environmental surveillance of typhoid and antimicrobial resistance through urban wastewater are emerging areas of interest in global and public health organizations \cite{Pehrsson2016,Initiatives2018}. Besides infectious disease, methods have been developed and standardized over the last decade to successfully provide continuous data on illicit drug consumption at the national- and regional-level in Australia \cite{AustralianReport} and over 60 European cities and towns \cite{vanNuijs2018,Thomas2012,BazLomba2016}. The academic community is now interested in branching out to measure lifestyle chemicals and biomarkers of health such as obesity levels \cite{Daughton2018,Arnaud2018,Newton2015}. More recently, wastewater testing has been proposed as a promising tool to tackle the opioid epidemic that is severely affecting the U.S \cite{Keshaviah2016,Keshaviah2017}. However, apart from successful poliovirus surveillance and, to a lesser extent, national-level drug consumption surveillance programs, most application areas remain proof-of-concept academic projects and have yet to be implemented into public health and other public service delivery agencies.

@@

Besides providing geographical specificity to the sample, residential catchments mitigate variability in population size and sewage travel time observed at WWTPs. Our selection criteria produces catchments with similar population sizes by design (5,000--10,000 inhabitants) rather than the large range observed (1,000--2,000,000 inhabitants) in WWTPs (Supp. Fig. \ref{24hr:figS1}, \cite{Newton2015}). Additionally, residential catchments are more likely to contain even representation of all households in the catchment compared to WWTPs: in our residential catchment, sewage from every source travels at most 45 minutes to the sampling point and the distribution of travel times is narrowly-distributed (Fig. \ref{24hr:fig1}A). Sewage at WWTPs, on the other hand, can have travelled anywhere from minutes to hours before it reaches the sampling point. Sampling in residential catchments therefore mitigates the need to correct for unequal population sizes and travel times to some extent.

@@

Metabolites and bacteria in sewage separate into human and environmental contributions. Groups M1 and O1 were enriched in human-derived bacteria and metabolites, while groups M2 and O2 contained bacteria and metabolites derived from the environment (Fig. \ref{24hr:fig3}C). The cluster M1 was statistically enriched in matches to human metabolites compared to the M2 cluster (Fig. \ref{24hr:fig3}C), and includes known urinary and fecal metabolites. Twenty-two metabolic features in M1 were identified as glucuronide compounds of hormones, bile acids and pharmaceuticals, which are direct markers of human excretion as discussed above. Bacteria in the cluster O1 were putatively identified as coming from human gut microbiomes. Cluster O1 included 41 bacterial OTUs primarily from the Firmicutes and Bacteroidetes phyla, which are major components of the human gut flora (Fig. \ref{24hr:fig3}D). Additionally, the top 10 most abundant bacterial families in human stool (Human Microbiome Project) made up 65\% of the community in the O1 group (Supp. Fig. \ref{24hr:figS4}). Similar to what was observed in the metabolites, O1 bacteria had a dip in relative abundance during nighttime (Figure \ref{24hr:fig3}B). Clusters M2 and O2, on the other hand, remained relatively constant through the day and likely reflect bacteria and metabolites sourced from environmental background rather than human activity. Cluster M2 had few putative human-associated hits, and had constant abundance through the day and nighttime. The molecules in this group likely represent chemicals derived from the environment or sewer biochemistry. Human-associated bacteria were similarly decreased in cluster O2: most of the OTUs were Proteobacteria, which are not major members of the normal gut flora, and human fecal bacterial families comprised only 1.4\% of the abundance of this cluster (Supp. Fig. \ref{24hr:figS4}). The relative abundance of cluster O2 increased at night, reflecting that these environmental made up a larger proportion of the residential sewage microbiome during periods of low human activity (Fig. \ref{24hr:fig3}D).
